---
title: "sdmTMB lab: index standardization"
format: html
editor: visual
---

In this document, we'll work through common modeling workflows for ecologists and fisheries scientists using `sdmTMB` to do index standardization.

Along the way, we'll discuss model validation and residual checks, results visualization, and explore how tweaking model parameters impacts outcomes.

### Packages and data

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(sdmTMB)
library(here)
library(INLA) #
library(gstat) #
library(mgcv) 
library(sp)
library(DHARMa) #
library(sdmTMBextra) #remotes::install_github("pbs-assess/sdmTMBextra", dependencies = TRUE)

# load processed data from R/data_processing_and_visualization
load(here::here("data/processed_hair_crab_data.rdata"))

# load hair crab data (includes design-based index)
load(here::here("data/hair_cpue.rdata"))

# source helper functions
source(here::here("R/helpers.R"))

# Convert sf to data.frame for use with sdmTMB
hair_cpue <- 
  hair_cpue_sf_filt %>% 
    sfc_as_cols(., names = c("longitude", "latitude")) %>% # turn points into lat/lon columns
    st_set_geometry(NULL) %>% #convert to data.frame
  filter(year >= 2014) %>% 
  mutate(district = factor(district))

# CRS
ak_crs <- "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +datum=NAD83 +units=km +no_defs"
```

### Mesh creation

#### Different approaches for mesh creation

`sdmTMB` internal functions will build a mesh for you if you'd like, although these functions still require user input and optimization. An alternative approach is to build the mesh yourself using functions from `INLA`. We will demonstrate the custom approach first so that you are familiar with the language associated with meshes.

#### Building a custom mesh with INLA

The `inla.mesh.2d` function is used to build the mesh, and selecting the parameters that govern mesh structure requires trial and error. The steps I describe here are based on those outlined in Zuur et al. 2017, Chapter 12 and Zuur et al. 2018, Chapter 19, <https://rpubs.com/jafet089/886687>, and <https://haakonbakkagit.github.io/btopic104.html>.

1.  Estimate a rough spatial range at which residual spatial autocorrelation diminishes. We do this by fitting a Tweedie GLM and then evaluating the distance among scaled residuals at which spatial correlations decline.

```{r, message=FALSE,warning=FALSE}
# fit a Tweedie GLM (ignore spatial autocorrelation)
tw_mod <- sdmTMB(cpue ~ 0 + as.factor(year), 
                 data = hair_cpue, 
                 spatial = "off",
                 family = tweedie())

# DHARMa residuals
resid_sim_tw_mod <- simulate(tw_mod, nsim = 500, type = "mle-mvn")
resid_dharm_tw_mod <- dharma_residuals(resid_sim_tw_mod, tw_mod, return_DHARMa = TRUE)
hair_cpue$tw_mod_resids <- resid_dharm_tw_mod$scaledResiduals

# Plot sample variogram for scaled residuals
gg_out <- NULL
for (i in unique(hair_cpue$year)){
  hair_cpue_yr <- hair_cpue  %>% 
    filter(year == i) %>% 
    as.data.frame()
  sp::coordinates(hair_cpue_yr) <- c("longitude","latitude") 
  v1 <- gstat::variogram(tw_mod_resids ~ 1, 
                      hair_cpue_yr)
  
  assign("gg_out", rbind(tibble(distance = v1$dist, 
                                Semivariance = v1$gamma,
                                year = i),
                         gg_out))
}

rough_range <- 100

ggplot(data = gg_out, aes(y = Semivariance, 
                                x = distance, color = year)) + 
  geom_point() +
  geom_smooth() +
  geom_vline(xintercept = rough_range) +
  labs(x = "Distance (km)") + 
  facet_wrap(~year, scales = "free_y")
```

Here we use sample variograms to visualize how residual spatial autocorrelation changes with distance. If residual spatial autocorrelation is isotropic, then we would expect to see semivariance consistently increase and plateau where residuals are no longer well correlated.

The deviations from our expectations about semivariance shown here may be related to the fact that hair crab are patchily distributed in the Eastern Bering Sea. For now, we will pick a spatial range of 100 km for developing the mesh.

2.  We next specify the spatial domain by passing sampling coordinates to the `loc.domain` argument. This creates a concave hull around the sampling locations.

3.  We then address mesh resolution using `max.edge`. This sets the maximum length of edges within the inner mesh. The maximum edge length for the outer boundary is set by adding a second value to this argument. Guidance from an INLA developer suggests that `max.edge` should be between 1/10 and 1/5 of the spatial range (<https://rpubs.com/jafet089/886687>) as a starting point. We will use a slightly coarser value in favor of estimation speed.

    The outer boundary serves the purpose of minimizing edge effects (large variance in vertex weights). The extent of the outer boundary is set by the `offset` argument. We set this value to be greater than `max.edge` to provide multiple "layers" of triangles in the outer domain.

    -   Once the spatial range is estimated be `sdmTMB`, we can update our initial range in the mesh.

```{r}
## INLA mesh----
max.edge <- rough_range/3
offset <- rough_range/2
custom_mesh <- INLA::inla.mesh.2d(loc.domain = st_coordinates(hair_cpue_sf_filt),
                           max.edge = c(max.edge, max.edge * 2),
                           offset = c(max.edge, offset))

# we need to convert the mesh prior to use in sdmTMB
custom_mesh_sdmTMB <- sdmTMB::make_mesh(data = hair_cpue,
                              xy_cols = c("longitude", "latitude"),
                              mesh = custom_mesh)

plot(ak_land, col = "darkgreen")
plot(custom_mesh, add = T)
points(hair_cpue$longitude, hair_cpue$latitude, col="blue")
```

4.  Haakon Bakka's third suggestion for "What is needed for a good mesh" is that there should be no "weird-looking" parts. One thing that is immediately obvious in this mesh is that it extends over land. Is this an issue?

    Probably not unless there is a very large island splitting the study domain. The model assumes correlations decay over Euclidean distances and therefore would "borrow strength" for predictions using observation nearby in Euclidean terms but much further away in terms of how crabs move. At the spatial scale of the Eastern Bering Sea, I would be surprised if this mattered. Nonetheless, here is a "barrier mesh" version of the previous mesh. We will come back to this to see if it matters.

```{r, message=FALSE}
## INLA mesh with barrier specified----
max.edge <- rough_range/3
offset <- rough_range/2
hair_loc_buffers_sp <- hair_loc_buffers %>% as_Spatial() #must be sp class, not sf
barrier_mesh <- INLA::inla.mesh.2d(boundary = hair_loc_buffers_sp,
                           max.edge = c(max.edge, max.edge * 2),
                           offset = c(max.edge * 2, offset))

# we need to convert the mesh prior to use in sdmTMB
barrier_mesh_sdmTMB <- sdmTMB::make_mesh(data = hair_cpue,
                                          xy_cols = c("longitude", "latitude"),
                                          mesh = barrier_mesh)

plot(ak_land, col = "darkgreen")
plot(barrier_mesh, add = T)
points(hair_cpue$longitude, hair_cpue$latitude, col="blue")
```

#### Allowing sdmTMB to make the mesh

The simpler option is to let internal functions within `sdmTMB` build the mesh for you using `make_mesh`. Within `make_mesh`, if `type = "cutoff"` then the user may adjust resolution by setting the `cutoff` value: lower cutoff provides a higher resolution mesh. If `type = "kmeans"` then the mesh vertices are assigned using an k-means algorithm, and the number of knots can be set with `n_knots.`

```{r}

# Approaches within sdmTMB vignette------
# Extremely simple cutoff:
sdmTMB_mesh1 <- make_mesh(hair_cpue, c("longitude", "latitude"), 
                          cutoff = 5, type = "cutoff")
plot(sdmTMB_mesh1)

# Using a k-means algorithm to assign vertices:
sdmTMB_mesh2 <- make_mesh(hair_cpue, c("longitude", "latitude"), 
                          n_knots = 150, type = "kmeans")
plot(sdmTMB_mesh2)
```

#### Mesh complexity

The number of vertices in the mesh is one of the primary factors determining computation time. To speed up modeling workflows, you might initially choose a coarser resolution mesh as you iterate on model structure.

```{r}
tibble(`N vertices` = c(custom_mesh_sdmTMB$mesh$n,
                        barrier_mesh_sdmTMB$mesh$n,
                        sdmTMB_mesh1$mesh$n,
                        sdmTMB_mesh2$mesh$n),
       Type = c("Custom - no barrier", 
                "Custom - barrier", #
                "sdmTMB - cutoff",
                "sdmTMB - kmeans"))
```

## Index standardization and comparisons with design-based indices

A common use of `sdmTMB` is to estimate the total or relative abundance or biomass of a species within a region over time, i.e., biomass or abundance "indices". `sdmTMB` contains functions that will generate these indices when given a fitted model. We will explore these below, but it's worth noting that these functions are simply taking model predictions and aggregating them in some way (e.g., summing predictions over space).

A benefit to using `sdmTMB` to generate these indices is that `sdmTMB` applies the *delta method* internally to calculate standard errors on predictions, regardless of user-selected aggregation function.

First, we need to fit a `sdmTMB` model. For simple index standardization, the relevant components are:

1.  `formula` - the standard R formula interface. In this case, we are asking the model to fit categorical, fixed effects intercepts for each year of data. We ensure that an intercept is produced for all years by suppressing the global model intercept with `~0`. We will add further complexity to this formula later.
2.  `spatial` - when `on`, the model estimates a spatial random field. This is a "persistent deviation" in density across the spatial domain.
3.  `spatiotemporal` - when specified, the model estimates a spatiotemporal random field. This is a random field that varies across time steps. We specify `iid` to start, meaning that the random field is estimated as independent through time. Spatiotemporal correlations are modeled when this is `ar1` or `rw`
4.  `time` - A string specifying the column name of the numeric time variable in input data frame.
5.  `mesh` - the mesh object used to estimate the spatial and spatiotemporal random fields. Not necessary if spatial terms are excluded.
6.  `family` - the distribution for observation error and the link function. Here is where we specify delta models if desired.

### Model fitting

Here we fit a model using Tweedie observation error with a log link, spatial, and spatiotemporal random fields (iid).

```{r}
process <- TRUE
if (process){
  m1 <- sdmTMB::sdmTMB(
          formula = cpue ~ 0 + as.factor(year),
          spatial = "on",
          spatiotemporal = "iid",
          time = "year",
          mesh = sdmTMB_mesh2, # simple mesh for speed
          family = tweedie(link = "log"),
          data = hair_cpue)
  save(m1, file = here::here("data/m1.rdata"))
} else {
  load(here::here("data/m1.rdata"))
}
```

### Model evaluation

#### Model summary

```{r}
m1
```

A few key elements here.

1.  The `dispersion parameter` and `Tweedie p` are parameters associated with the Tweedie distribution estimated from the data.
2.  The `Matern range` specifies the distance in km at which observations are no longer substantially correlated (if coordinates are in units of km).
3.  `Spatial SD` and `Spatiotemporal IID SD` - Marginal standard deviations of the spatial and spatiotemporal random fields. How variable is the spatial random field, and how variable are the spatiotemporal random fields?
4.  `ML criterion at convergence` - The maximized joint negative log likelihood for the fitted model.

#### Model diagnostics

Identifying issues with fitted models is made easy with the `sanity` function. This function also runs internally when the model fits in the `sdmTMB` function, and will produce a message at the end if an issue occurs.

```{r, eval=TRUE}
sanity(m1) #Tweedie model
```

We are met with issues from this fitted model, including `ln_kappa gradient > 0.001` and `ln_phi gradient > 0.001`. The kappa being referred to is either the spatial or spatiotemporal decorrelation rate, and phi is the Tweedie dispersion parameter.

Our initial analysis of residual spatial correlation points to the likely issue and how to solve it. `sdmTMB` models assume isotropic stationarity in spatial correlations (i.e., correlation is independent of direction). We can relax this assumption and allow for directionally dependent spatial correlations by setting `anisotropy = TRUE`.

```{r}
if (process){
  m1.2 <- sdmTMB::sdmTMB(
          formula = cpue ~ 0 + as.factor(year),
          spatial = "on",
          spatiotemporal = "iid",
          time = "year",
          anisotropy = TRUE,
          mesh = sdmTMB_mesh2,  # simple mesh for speed
          family = tweedie(link = "log"),
          data = hair_cpue)
  save(m1.2, file = here::here("data/m1_2.rdata"))
} else {
  load(here::here("data/m1_2.rdata"))
}
m1.2
```

This addition resolves the errors in `sanity`. `sdmTMB` also include functions to visualize the directionality of anisotropy. The major axis of spatial correlation aligns with the distribution of hair crab in Bristol Bay, and the minor axis aligns with the relationship between the Bristol Bay and Pribilof Islands stocks.

```{r}
plot_anisotropy(m1.2)
plot_anisotropy2(m1.2)
```

#### Evaluating model residuals

There are a few ways to evaluate model residuals in `sdmTMB`. Detailed examples are provided [here](https://sdmtmb.github.io/sdmTMB/articles/residual-checking.html). The main three are as follows:

1.  **Randomized quantile residuals** - Residuals are transformed from Tweedie (in this case) to a uniform distribution (cumulative probabilities). Probabilities are transformed to normal quantiles. Well-behaved rediuals will appear normal.
    -   Residuals calculated by sampling once from the approximate MVN distribution of the random effects, conditional on fixed effects held at their MLEs.
    -   Best to set a seed to avoid fishing for good looking residuals.

```{r}
set.seed(67)
r1 <- residuals(m1.2, type = "mle-mvn")
qqnorm(r1);abline(0, 1)
ks.test(r1, pnorm) # test to see if residuals deviate from normal
```

2.  **Simulation-based randomized quantile residuals** (DHARMa)

```{r, message=FALSE}
set.seed(67)
r2 <- simulate(m1.2, nsim = 1000, type = "mle-mvn") %>% #simulate
  dharma_residuals(m1.2, return_DHARMa = TRUE) # pass to DHARMa

plot(r2)
```

3.  **MCMC based residuals** - Relaxes assumptions of multivariate normality for random effects for randomized quantile residuals. This approach uses MCMC to sample the REs with fixed effects held at their MLEs.

```{r}
if (process){
  samps <- sdmTMBextra::predict_mle_mcmc(m1.2, mcmc_iter = 2000, mcmc_warmup = 800)
  mcmc_res <- residuals(m1.2, type = "mle-mcmc", mcmc_samples = samps)
  save(mcmc_res, file = here::here("data/m1_mcmc_resids.rdata"))
} else {
  load(here::here("data/m1_mcmc_resids.rdata"))
}
qqnorm(mcmc_res)
abline(0, 1)
ks.test(mcmc_res, pnorm)
```

MCMC based residuals are more robust than randomized quantile residuals due to relaxation of assumptions on REs, although take substantially more time to generate. DHARMa residuals are useful because you get easy access to the various tests available in DHARMa.

We identified some departure from residual expectations for this model when applying the analytical and simulation-based quantile residuals approach. The MCMC-based residuals behave as expected, although warnings during sampling suggest that MCMC chains may not have mixed well. Troubleshooting thing will take iteration, but it is worth exploring all three approaches to stress test your models.

### Predicting from the model

Predicting from `sdmTMB` models is fun and rewarding. Here we will generate interannual predictions of hair crab density over a gridded spatial domain (the prediction grid) and compare them visually with observed CPUE. We need to expand our prediction grid so that we are generating predictions for each year.

#### Generating the prediction grid

The general workflow for predicting from sdmTMB models at new locations and/or times is as follows:

1.  Select a grid cell size and create the grid. In the following I provide you with the workflows to either a) alter grid cell resolution to use your own, or b) you can instead choose the gridded region used in the development of design-based indices for the EBS.
2.  Expand the data so that all grid cells are available at all time steps. Additional covariates must be included as columns.
3.  Extract grid cell centroids and convert to a data frame with columns for your longitude and latitude variables.

```{r, message=FALSE, warning=FALSE}
custom_resolution <- TRUE
# Set custom grid cell resolution
if (custom_resolution){
  cellsize <- 20 #km x cellsize km

  # polygon for the EBS
  ebs_domain <- 
    st_read(here::here("data/SAP_layers.gdb"),
          layer = "EBS_grid",
          quiet = TRUE) %>% 
    st_transform(., ak_crs) %>% 
    st_union()
  
  # re-grid it to our spatial resolution
  ebs_grid <- ebs_domain %>% 
    st_make_grid(., cellsize  = cellsize) %>% 
    st_intersection(., hair_loc_buffers) %>%
    st_as_sf() %>% 
    mutate(id = 1:nrow(.),
           area = as.numeric(st_area(.)))
  
  # expand over time
  pred_grid <- ebs_grid %>% 
    tidyr::expand_grid(.,
                       year = min(unique(hair_cpue$year)):max(unique(hair_cpue$year))) %>% 
    st_as_sf() %>% 
    filter(year != 2020) # COVID year - no survey so should not be in final data
  
  # final outputs
  pred_df <- pred_grid %>% 
    st_centroid() %>% 
    sfc_as_cols(., names = c("longitude","latitude")) %>% 
    st_set_geometry(NULL)
  
} else {
    # polygon for the EBS
  ebs_grid <- 
    st_read(here::here("data/SAP_layers.gdb"),
          layer = "EBS_grid",
          quiet = TRUE) %>% 
    st_transform(., ak_crs) %>% 
    st_intersection(., hair_loc_buffers) %>%
    mutate(area = as.numeric(st_area(.))) %>% 
    dplyr::select(area) %>% 
    mutate(id = 1:nrow(.))
  
  # expand over time
  pred_grid <- ebs_grid %>% 
    tidyr::expand_grid(.,
                       year = min(unique(hair_cpue$year)):max(unique(hair_cpue$year)),
                       var = "hair_crab") %>% 
    st_as_sf() %>% 
    filter(year != 2020) # COVID year - no survey so should not be in final data
  
  # convert to df
  pred_df <- pred_grid %>% 
    st_centroid() %>% 
    sfc_as_cols(., names = c("longitude","latitude")) %>% 
    st_set_geometry(NULL) %>% 
    as.data.frame()
}

```

#### Model predictions

Our goal is to generate an abundance index that approximates the design-based index for legal male hair crab. The model-based index in the NMFS EBS survey report is produced for the full Eastern Bering Sea. Therefore, there is no need to filter our prediction data frame prior to using it as `newdata` in `sdmTMB::predict` - we've already trimmed the prediction grid to the area of the EBS that overlaps with where hair crab occur. `return_tmb_object = TRUE` is required for passing predictions to the `get_index` function for index calculation.

```{r}
model_predictions_df <- predict(m1.2, newdata = pred_df, type = "link",
                                return_tmb_object = TRUE) 
model_predictions_df$data
```

The prediction data frame contains several columns of interests:

1.  `est` - Predicted hair crab log-density (set `type = response` for the estimate on the response scale).
2.  `est_non_rf` - Log-density associated with fixed effects (year).
3.  `est_rf` - Log-density associated with random effects (spatial and spatiotemporal random fields).
4.  `omega_s` - Log-density associated with spatial random field.
5.  `epsilon_st` - Log-density associated with spatiotemporal random field.

Visualizing predictions by year:

```{r, fig.width=8.5, fig.height=11}
model_predictions_sf <- pred_grid %>% 
  left_join(., model_predictions_df$data, by = c("id","year")) 

ggplot() +
  geom_sf(data = ak_land, fill = "darkgreen", alpha = 0.75) +
  geom_sf(data = model_predictions_sf, aes(fill = est, color = est)) +
  geom_sf(data = hair_cpue_sf_filt %>% filter(cpue > 0,
                                              year >= 2014),
          aes(size = log(cpue)), color = "red", alpha = 0.5) +
  scale_fill_gradient2(low = scales::muted("red"),
                       high = scales::muted("blue"))+
    scale_color_gradient2(low = scales::muted("red"),
                       high = scales::muted("blue"))+
  guides(color = "none") +
  labs(fill = "Predicted log(CPUE)",
       size = "Observed log(CPUE>0)") +
  theme_light() + 
  coord_sf(ylim = c(650, 1500),
         xlim = c(-1300, -175)) +
  facet_wrap(~year, ncol = 3) +
  theme(legend.position = "bottom")
```

### Generating the abundance index

Generating the final index is quite easy using the `get_index` function. Remember that the `area` argument is in sq. km. In the following script, we calculate the model-based index using `get_index`, and bind that to the design-based indices produced by `crabpack`.

```{r, fig.height=5, fig.width=8}
model_based_index <- 
  get_index(obj = model_predictions_df,
          bias_correct = TRUE,
          area = pred_df$area) %>% 
  mutate(type = "Model-based")

design_based_index <- 
  hc_design_index_ebs %>% 
    rename_all(str_to_lower) %>% 
    mutate(lwr = ifelse((abundance - abundance_ci) < 0, 0,
                                (abundance - abundance_ci)),
           upr = abundance + abundance_ci,
           type = "Design-based") %>% 
    dplyr::select(year, 
                  est = abundance,
                  lwr,
                  upr,
                  type) %>% 
  filter(year >= 2014)

combined_indices <- model_based_index %>% 
  bind_rows(., design_based_index)


ggplot(combined_indices)+
  geom_errorbar(aes(ymax = upr, ymin = lwr, x = year, color = type), width = 0, linewidth = 1) +
  geom_line(aes(y = est, x = year, color = type), linewidth = 1) +
  scale_color_manual(values = c("Model-based" = "darkorange","Design-based" = "purple")) +
    labs(title = "Design-based vs. model-based hair crab abundance indices",
       y = "Abundance",
       x = "Year") +
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 12),
        axis.title =  element_text(size = 12),
        title  = element_text(size = 14))
```

The design-based and model-based indices are roughly the same, as expected

What if we're looking for an index on the scale of the units? For instance, if our observations are catch per pot or number per unit hour etc. We can simulate predictions from the joint precision matrix and apply any aggregation function over those draws.

```{r}
model_predictions_df_sim <- predict(m1.2, 
                                    newdata = pred_df, 
                                    nsim = 1000,
                                    type = "link")

model_based_index_relative <- 
  get_index_sims(obj = model_predictions_df_sim,
                 agg_function = function(x) median(exp(x)),
                 est_function = median)

ggplot(model_based_index_relative)+
  geom_errorbar(aes(ymax = upr, ymin = lwr, 
                    x = year), width = 0, linewidth = 1) +
  geom_line(aes(y = est, x = year), linewidth = 1) +
    labs(y = "Catch km^-2",
         x = "Year",
         title = "Simulation-based relative index") +
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 12),
        axis.title =  element_text(size = 12))
```

Keep in mind that we only fitted a single model in this analysis. Next, we will discuss model comparison and selection.
