---
title: "sdmTMB lab: data pre-processing and visualization"
format: html
editor: visual
---

### Packages and data

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(mgcv)

# load hair crab data
load(here::here("data/hair_cpue.rdata"))

# Coordinate reference system for projecting CPUE data (Albers Equal Area)
# Ensure units are km
ak_crs <- "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +datum=NAD83 +units=km +no_defs"

# convert data frame to sf object and project
hair_cpue_sf_all <- hair_cpue %>% 
  st_as_sf(., coords = c("longitude","latitude"), crs = 4326) %>% # make sf
  st_transform(., crs = ak_crs)

bb_poly <- st_as_sfc(st_bbox(c(
      xmin = -180,  xmax = -156,
      ymin =  50, ymax =  70), 
    crs = 4326))

# Land polygon
ak_land <- rnaturalearthhires::states10 %>%
  filter(name == "Alaska") %>%
  st_union() %>% 
  st_intersection(bb_poly) %>% 
  st_transform(ak_crs)
```

### Visualize survey effort distribution

```{r}
ggplot() + 
  geom_sf(data = hair_cpue_sf_all, 
          aes(color = district), 
          size = 1) +
  geom_sf(data = ak_land) +
    coord_sf(ylim = c(550, 1800),
           xlim = c(-1300, -175)) +
  facet_wrap(~year)
```

#### Notes

-   Northern Bering Sea (NBS) is not surveyed every year like the Eastern Bering Sea (EBS)
-   EBS data prior to 2024 does includes "corner stations" around Pribilof Islands and St. Matthews Island
-   Survey stations are not identical across years in NBS - sampling effort is more sparse in 2018

### Visualize hair crab distribution

```{r}

# just look at >2011 data
cpue_dist <- 
  hair_cpue_sf_all %>% 
  filter(cpue > 0, year > 2011)

ggplot() + 
  geom_sf(data = ak_land) +
  geom_sf(data = cpue_dist, aes(color = cpue),
          size = 0.5) +
  geom_sf(data = cpue_dist, color = "red", size = 2) +
  coord_sf(ylim = c(550, 1800),
           xlim = c(-1300, -175)) +
  facet_wrap(~year)
```

### Identify regions where hair crab are absent from all trawls

-   To aid in model estimation, we will filter out zero observations outside a 45 km distance from observed positive catch values (filtering "false zeros")
-   The blue polygon below indicates the region within which we will predict hair crab densities.

```{r, message=FALSE, warning=FALSE}

hair_loc_buffers <- hair_cpue_sf_all %>% # across all trawls
  filter(cpue > 0) %>% #
  concaveman::concaveman(., concavity = 3) %>% #Draw concave hull around positive catches
  st_buffer(., dist = 45) %>%  # extend buffer
  st_difference(ak_land) #remove sections of polygon intersecting land

ggplot() +
  geom_sf(data = hair_loc_buffers, fill = "lightblue") +
  geom_sf(data = hair_cpue_sf_all %>% filter(cpue == 0), color = "red") +
  geom_sf(data = hair_cpue_sf_all %>% filter(cpue > 0))

```

### Intersect buffer polygon with survey data

-   These will be the data that we use to fit the model.
-   To simplify analyses, we combine the `district` variable "ALL" and "NORTH" into a general "N. EBS/NBS" district.

```{r, message=FALSE, warning=FALSE}
hair_cpue_sf_filt <- hair_cpue_sf_all %>%
  mutate(district = ifelse(district %in% c("ALL", "NORTH"), 
                           "N. EBS/NBS",
                           district)) %>% 
  st_intersection(., hair_loc_buffers)
```

#### Notes

-   Positive densities range from SE Bristol Bay into the SE NBS
-   Positive catches also occur in the Pribs and St. Matt's
-   Rare occurrences west of Pribs and in W. NBS

### Visualize relationships with environmental covariates

#### Temperature (gear temperature)

```{r, message=FALSE, warning=FALSE}
ggplot(hair_cpue_sf_filt) + 
  geom_point(aes(y = cpue, x = temp))+
  facet_wrap(~district)
```

#### Depth (bottom depth)

```{r, message=FALSE, warning=FALSE}
ggplot(hair_cpue_sf_filt) + 
  geom_point(aes(y = cpue, x = bottom_depth)) +
  facet_wrap(~district)
```

### Export processed data

```{r, message=FALSE, warning=FALSE}
save(hair_cpue_sf_filt,
     hair_cpue_sf_all,
     hair_loc_buffers,
     ak_land,
     file = here::here("data/processed_hair_crab_data.rdata"))
```
