---
title: "sdmTMB lab: model comparison and selection"
format: html
editor: visual
---

In this document, we'll step through model comparison workflows in `sdmTMB`. We'll discuss including smooth effects and how to visualize them, a couple of approaches for comparing models, and also how to do Bayesian inference with `sdmTMB` and `tmbstan`.

### Packages and data

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(sdmTMB)
library(here)
library(INLA) #
library(gstat) #
library(mgcv) 
library(sp)
library(DHARMa) #
library(ggeffects)
library(tmbstan)
library(sdmTMBextra) #remotes::install_github("pbs-assess/sdmTMBextra", dependencies = TRUE)

# load processed data from R/data_processing_and_visualization
load(here::here("data/processed_hair_crab_data.rdata"))

# load hair crab data (includes design-based index)
load(here::here("data/hair_cpue.rdata"))

# source helper functions
source(here::here("R/helpers.R"))

# Convert sf to data.frame for use with sdmTMB
hair_cpue <- 
  hair_cpue_sf_filt %>% 
    sfc_as_cols(., names = c("longitude", "latitude")) %>% # turn points into lat/lon columns
    st_set_geometry(NULL) %>% #convert to data.frame
  filter(year >= 2014) %>% 
  mutate(district = factor(district),
         yf = factor(year),
         depth = as.numeric(scale(bottom_depth)))

# CRS
ak_crs <- "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +datum=NAD83 +units=km +no_defs"

# mesh used in index standardization analysis
sdmTMB_mesh2 <- make_mesh(hair_cpue, c("longitude", "latitude"), 
                          n_knots = 150, type = "kmeans")
```

### Model fitting

In the previous analysis we identified a well-fitting model that was effective for the purpose of developing an abundance index. However, we did not evaluate the goodness of fit of that model relative to other potential models.

We will explore how to compare models here. Imagine that instead of index standardization we were specifically interested in understanding the preferred depth of hair crab.

We model hair crab CPUE as a smooth function of depth, in addition to the inclusion of a spatial random field, a random effect for year, and anisotropy.

We compare this model with one that also includes estimates district-specific ("N EBS/NBS", "BB", and "PRIB") intercepts.

To make the models fit, we include `m = 1` in the smooth term, where `m` is the order of penalized derivatives (usually m = 2 for TPRS). In this case, setting m = 1 has the practical impact of preventing smooth predictions of CPUE from having enormous uncertainty for the deepest bottom depths. Additionally, we limit the number of basis functions included in `m2` to 5. This simplification prevents `sanity` errors.

```{r}
m2 <- sdmTMB::sdmTMB(
        formula = cpue ~ (1|yf) + s(bottom_depth, m = 1, k = 5),
        spatial = "on",
        anisotropy = TRUE,
        mesh = sdmTMB_mesh2,
        family = tweedie(link = "log"),
    data = hair_cpue)

# sanity check
sanity(m2)

set.seed(67)
r2 <- residuals(m2, type = "mle-mvn")
qqnorm(r2);abline(0, 1)
```

We might suspect that a better model would include varying intercepts by district,

```{r}
m3 <- sdmTMB::sdmTMB(
        formula = cpue ~ (1|yf) + s(bottom_depth, m = 1) + district,
        spatial = "on",
        anisotropy = TRUE,
        mesh = sdmTMB_mesh2,
        family = tweedie(link = "log"),
    data = hair_cpue)

# sanity check
sanity(m3)

set.seed(67)
r3 <- residuals(m3, type = "mle-mvn")
qqnorm(r3);abline(0, 1)
```

Both models pass sanity checks. The model `m3` has slightly uglier residuals on the qq-plot.

Let's examine what the marginal smooths for depth look like prior to model comparison.

### Visualizing smooth effects

```{r}
ggpredict(m2, terms = c("bottom_depth")) %>% plot()
ggpredict(m3, terms = c("bottom_depth", "district")) %>% plot()
```

### Model comparisons

Our first comparison is to look at conditional AICs between these two models. ?sdmTMB::cAIC states "cAIC is designed to optimize the expected out-of-sample predictive performance for new data that share the same random effects as the in-sample (fitted) data, e.g., spatial interpolation. In this sense, it should be a fast approximation to optimizing the model structure based on k-fold cross-validation."

```{r}
cAIC(m2)
cAIC(m3)

AIC(m2, m3)
```

cAIC and AIC values support the inclusion of `district` for improving interpolation (cAIC) and extrapolation/inference performance (AIC).

`sdmTMB` provides several cross-validation methods that allow for comparing model prediction skill. The out-of-the-box method drops a chunk (i.e. a "fold") of data, re-fits the model without it, and predicts on the left out data. You can specify the number of folds with `k_folds`.

```{r}
process <- TRUE
if (process){
  m_2_cv <- 
    sdmTMB::sdmTMB_cv(
          k_folds = 8,
          formula = cpue ~ (1|yf) + s(bottom_depth, m = 1, k = 5),
          spatial = "on",
          anisotropy = TRUE,
          mesh = sdmTMB_mesh2,
          family = tweedie(link = "log"),
          data = hair_cpue)
  
  m_3_cv <- 
   sdmTMB::sdmTMB_cv(
          k_folds = 8,
          formula = cpue ~ (1|yf) + s(bottom_depth, m = 1) + district,
          spatial = "on",
          anisotropy = TRUE,
          mesh = sdmTMB_mesh2,
          family = tweedie(link = "log"),
          data = hair_cpue)
  
  save(m_2_cv, m_3_cv, file = here::here("data/model_cross_val_results.rdata"))
} else {
  load(here::here("data/model_cross_val_results.rdata"))
}


m_2_cv$sum_loglik
mean(abs(m_2_cv$data$cpue - m_2_cv$data$cv_predicted)) #mean absolute error

m_3_cv$sum_loglik # more positive is better
mean(abs(m_3_cv$data$cpue - m_3_cv$data$cv_predicted)) #mean absolute error
```

Cross validation suggests there is very little additional predictive skill to be had through the inclusion of `district`

### Bayesian inference on parameters

The most straightforward way to evaluate hypotheses for sdmTMB models is to fit the model in a Bayesian context. To do this, we simply flag `bayesian = TRUE` in the sdmTMB model fit and pass the model object to `tmbstan`.

In this case, we did not specify priors for the model so they are implied uniform. Visit https://sdmtmb.github.io/sdmTMB/articles/bayesian.html for much more detail on fitting `sdmTMB` models in `tmbstan`.

```{r}
m3_b <- sdmTMB::sdmTMB(
          formula = cpue ~ 0 + (1|yf) + s(bottom_depth, m = 1) + district,
          spatial = "on",
          # anisotropy = TRUE,
          mesh = sdmTMB_mesh2,
          family = tweedie(link = "log"),
          bayesian = TRUE,
          data = hair_cpue)

tidy(m3_b)
tidy(m3_b, effects = "ran_vals")

# sanity check
sanity(m3_b)

fit_stan <- tmbstan::tmbstan(
  m3_b$tmb_obj,
  control=list(max_treedepth=12),
  iter = 1500, 
  chains = 4,
  seed = 6551
)

plot(fit_stan)
```
