---
title: "tinyVAST lab: index standardization and multivariate VAST models"
format: html
editor: visual
---

### Packages and data

The main objective of this lab is to explore common use-cases of `tinyVAST` for ecologists and fisheries scientists. We'll begin by porting over our index standardization model from sdmTMB to understand the components of tinyVAST. We'll then extend this univariate spatiotemporal model in a multivariate and vector autoregressive context. Lastly, we'll allow for the multivariate model to estimate two different response distributions for CPUE observations.

```{r,warning=FALSE,message=FALSE}
library(tidyverse)
library(sf)
library(tinyVAST)
library(sdmTMB)
library(here)
library(DHARMa)

# load processed data from R/data_processing_and_visualization
load(here::here("data/processed_hair_crab_data.rdata"))

# load hair crab data
load(here::here("data/hair_cpue.rdata"))

# source helper functions
source(here::here("R/helpers.R"))

# Convert sf to data.frame for use with sdmTMB
hair_cpue <- 
  hair_cpue_sf_filt %>% 
    sfc_as_cols(., names = c("longitude", "latitude")) %>% # turn points into lat/lon columns
    st_set_geometry(NULL) %>% #convert to data.frame
  filter(year >= 2014) %>% 
  mutate(district = ifelse(district == "N. EBS/NBS", "NEBS_NBS", district),
         var = "hair_crab",
         ) %>% 
  as.data.frame()

# CRS
ak_crs <- "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +datum=NAD83 +units=km +no_defs"

# load fitted sdmTMB model
load(here::here("data/m1_2.rdata"))

# We will use the same mesh for tinyVAST
sdmTMB_mesh2 <- make_mesh(hair_cpue, c("longitude", "latitude"), 
                          n_knots = 150, type = "kmeans")
tinyVAST_mesh <- sdmTMB_mesh2$mesh
```

#### Fit the equivalent tinyVAST model

All `tinyVAST` models require the user to specify arrow or arrow and lag notation expressing spatial, temporal, and/or spatiotemporal interactions among variables. Arrow and lag notation goes like this:

-   First term - Specifies the variables involved. `tinyVAST` works in long format. If we have two response variables X and Y, then we would indicate that grouping in a column variable containing levels for "X" and "Y". By default `tinyVAST` assumes this variable is called "var".

    -   In this first term we indicate the type of effect: One-headed arrow = slope parameter, two-headed arrow = variance/covariance

-   Second term - If this for a spatial interaction then the second term is the parameter name. If it's a temporal/spatiotemporal interaction, then the order of the lag.

-   Third term - For spatial interaction, this is an optional starting value for the parameter. For temporal/spatiotemporal interaction, this is the parameter name.

    -   When this term is set to "NA", then the parameter can be preset to a fixed value specified in the fourth term.

-   Fourth term - Only applicable when lags are specified - this is an optional starting value. If "NA" is specified, then the value of this term is fixed. For random walks, this would be 1.

Our `sdmTMB` model estimates a spatial random field and an i.i.d. spatiotemporal random field, and we need to code these into arrow notation prior to fitting the model. If no mesh is included (`spatial_domain = NULL`), then the model is treated as a time series model.

```{r}
# spatial random field
space_term <- "
  hair_crab <-> hair_crab, sd_space
"

spacetime_term <- "
  hair_crab <-> hair_crab, 0, sd_spacetime
"

m1_tv <- tinyVAST(
  space_term = space_term,
  spacetime_term = spacetime_term,
  data = hair_cpue,
  formula = cpue ~ 0 + as.factor(year),
  spatial_domain = tinyVAST_mesh,
  family = tweedie(),
  time_column = "year",
  space_columns = c("longitude","latitude"),
  variable_column = "var",
  control = tinyVASTcontrol(
    use_anisotropy = TRUE,
    newton_loops = 1 # sdmTMB sets newton_loops = 1 - set to 1 to get a similar sanity check
  )
)
m1_tv
sanity_tv(m1_tv)
```

### Comparisons with sdmTMB

#### Annual intercepts

```{r}
# extract parameters estimated as fixed effects (year intercepts)
year_seq <- c(2014:2019, 2021:2024)

# for tinyVAST, these are returned as "alpha_j"
tv_params <- m1_tv$sdrep
tv_names <- names(tv_params$par.fixed)
tv_year_intercepts <- tv_params$par.fixed[which(tv_names == "alpha_j")]

# for sdmTMB, these are "b_j"
sdmTMB_params <- m1.2$sd_report
sdmTMB_names <- names(sdmTMB_params$par.fixed)
sdmTMB_year_intercepts <- sdmTMB_params$par.fixed[which(sdmTMB_names == "b_j")]

tibble(tinyVAST_years = tv_year_intercepts,
       sdmTMB_year = sdmTMB_year_intercepts,
       year = year_seq)
```

`tinyVAST` and `sdmTMB` differ slightly in reporting style: `sdmTMB` provides more "plain English" descriptions of estimated parameters, and `tinyVAST` gives you variable names as reported by TMB. Ultimately these models are equivalent as expected.

#### Spatial random field

`theta_z`s contain your spatial interaction effects and spatial random field SDs

```{r}
tibble(tinyvast_ssd = tv_params$par.fixed[which(tv_names == "theta_z")],# tinyVAST spatial SD 
       sdmTMB_ssd = sdmTMB_params$value[1])  # sdmTMB spatial SD
```

#### Spatiotemporal random field

`beta_z`s contain spatiotemporal interaction effects and spatiotemporal random field SDs.

```{r}
tibble(tinyvast_st_sd = tv_params$par.fixed[which(tv_names == "beta_z")],# tinyVAST spatial SD 
       sdmTMB_st_sd = sdmTMB_params$value[3])  # sdmTMB spatial SD
```

#### Extend the model to model AR1 and random walk spatiotemporal correlations

We model a spatiotemporal correlation coefficient, rho, by specifying `hair_crab -> hair_crab, 1, rho` in the `spacetime_term` arrow and lag notation.

```{r}
space_term <- "
  hair_crab <-> hair_crab, sd_space
"

spacetime_term_ar1 <- "
  hair_crab -> hair_crab, 1, rho 
  hair_crab <-> hair_crab, 0, sd_spacetime
"

# AR1------
m1_tv_ar1 <- tinyVAST(
  space_term = space_term,
  spacetime_term = spacetime_term_ar1,
  data = hair_cpue,
  formula = cpue ~ 0 + as.factor(year),
  spatial_domain = tinyVAST_mesh,
  family = tweedie(),
  time_column = "year",
  space_columns = c("longitude","latitude"),
  variable_column = "var",
  control = tinyVASTcontrol(
    use_anisotropy = TRUE
  )
)
m1_tv_ar1

# Random walk------
spacetime_term_rw <- "
  hair_crab -> hair_crab, 1, NA, 1 
"

m1_tv_rw <- tinyVAST(
  space_term = space_term,
  spacetime_term = spacetime_term_rw,
  data = hair_cpue,
  formula = cpue ~ 0 + as.factor(year),
  spatial_domain = tinyVAST_mesh,
  family = tweedie(),
  time_column = "year",
  space_columns = c("longitude","latitude"),
  variable_column = "var",
  control = tinyVASTcontrol(
    use_anisotropy = TRUE
  )
)
m1_tv_rw

AIC(m1_tv, m1_tv_ar1, m1_tv_rw)
```

### Generate an annual index

#### Generate the prediction grid

```{r}
# polygon for the EBS
cellsize <- 20 #km x 20 km

# polygon for the EBS
ebs_domain <- 
  st_read(here::here("data/SAP_layers.gdb"),
        layer = "EBS_grid",
        quiet = TRUE) %>% 
  st_transform(., ak_crs) %>% 
  st_union()

# re-grid it to our spatial resolution
ebs_grid <- ebs_domain %>% 
  st_make_grid(., cellsize  = cellsize) %>% 
  st_intersection(., hair_loc_buffers) %>%
  st_as_sf() %>% 
  mutate(id = 1:nrow(.),
         area = as.numeric(st_area(.)))

# area vector to pass to integrate_output()
area_vec <- ebs_grid$area

# expand over time
pred_grid <- ebs_grid %>% 
  tidyr::expand_grid(.,
                     year = min(unique(hair_cpue$year)):max(unique(hair_cpue$year))) %>% 
  st_as_sf() %>% 
  filter(year != 2020) # COVID year - no survey so should not be in final data

# final outputs
pred_df <- pred_grid %>% 
  st_centroid() %>% 
  sfc_as_cols(., names = c("longitude","latitude")) %>% 
  st_set_geometry(NULL) %>% 
  as.data.frame()
```

#### Abundance index comparisons

The approach for generating abundance indices with `tinyVAST` is largely equivalent to `sdmTMB` using `tinyVAST::integrate_output()`. However, `tinyVAST` has additional options for weighting index calculations (using the `type` argument). We will use the default here that is equivalent to `sdmTMB`.

```{r}
# Abundance index calculations-----

## IID tinyVAST-----
est_iid <- sapply( year_seq ,
   FUN=\(t) integrate_output(m1_tv, newdata=subset(pred_df,year==t),
                             bias.correct = TRUE, area = area_vec) ) %>%
  t() %>% 
  as.data.frame() %>% 
  mutate(year = year_seq) %>% 
  dplyr::rename(est = Estimate,
                se = 2) %>% 
  mutate(lwr = est - 1.96 * se,
         upr = est + 1.96 * se,
         model = "tinyVAST - iid") %>% 
  dplyr::select(lwr, upr, model, year, est)

## AR1 tinyVAST-----
est_ar1 <- sapply( year_seq ,
   FUN=\(t) integrate_output(m1_tv_ar1, newdata=subset(pred_df,year==t),
                             bias.correct = TRUE, area = area_vec) ) %>%
  t() %>% 
  as.data.frame() %>% 
  mutate(year = year_seq) %>% 
  dplyr::rename(est = Estimate,
                se = 2) %>% 
  mutate(lwr = est - 1.96 * se,
         upr = est + 1.96 * se,
         model = "tinyVAST - AR1") %>% 
  dplyr::select(lwr, upr, model, year, est)

## RW tinyVAST-----
est_rw <- sapply( year_seq ,
   FUN=\(t) integrate_output(m1_tv_rw, newdata=subset(pred_df,year==t),
                             bias.correct = TRUE, area = area_vec) ) %>%
  t() %>% 
  as.data.frame() %>% 
  mutate(year = year_seq) %>% 
  dplyr::rename(est = Estimate,
                se = 2) %>% 
  mutate(lwr = est - 1.96 * se,
         upr = est + 1.96 * se,
         model = "tinyVAST - RW") %>% 
  dplyr::select(lwr, upr, model, year, est)

## IID sdmTMB-----
model_predictions_df <- predict(m1.2, newdata = pred_df, type = "link",
                                return_tmb_object = TRUE)
est_iid_sdmTMB <- get_index(obj = model_predictions_df, bias_correct = TRUE, area = pred_df$area) %>% 
  mutate(model = "sdmTMB - iid")

## Design-based-----
design_based_index <- 
  hc_design_index_ebs %>% 
    rename_all(str_to_lower) %>% 
    mutate(lwr = ifelse((abundance - abundance_ci) < 0, 0,
                                (abundance - abundance_ci)),
           upr = abundance + abundance_ci,
           model = "Design-based") %>% 
    dplyr::select(year, 
                  est = abundance,
                  lwr,
                  upr,
                  model) %>% 
  filter(year >= 2014)

combined_indices <- bind_rows(
  est_iid,
  est_ar1,
  est_rw,
  est_iid_sdmTMB,
  design_based_index
)

ggplot(combined_indices)+
  geom_errorbar(aes(ymax = upr, ymin = lwr, 
                    x = year, color = model), width = 0) +
  geom_line(aes(y = est, x = year, color = model)) +
    labs(title = "Design-based vs. model-based hair crab abundance indices",
       y = "Abundance",
       x = "Year") +
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 12),
        axis.title =  element_text(size = 12),
        title  = element_text(size = 14))
```

### Bivariate vector autoregressive model

A defining feature of `tinyVAST` is the ability to fit multivariate vector autoregressive models, where response variables are relate via cross-lagged effects.

In this example, we will fit a bivariate vector autoregressive spatiotemporal model evaluating cross-lagged effects of male hair crab density on female hair crab density, and vice versa.

#### Data processing

To begin, we repeat the same data processing workflow but now consider both male and female CPUE. Importantly, we arrange the data in long format, so that the `var` variable demarcates male and female CPUE groupings.

```{r}
start_year <- 2010

# combine male and female data and convert to sf
hair_cpue_sf_mf <- 
  hair_cpue_sf_all %>% 
  bind_rows(., hair_cpue_female %>% 
              st_as_sf(., coords = c("longitude", "latitude"), crs = 4326) %>% 
              st_transform(ak_crs)) %>% 
  dplyr::rename(var = category) %>% 
  filter(year >= start_year)

# create buffer around positive catches to drop false zeros before fitting model
hair_loc_buffers_mf <- hair_cpue_sf_mf %>% # across all trawls
  filter(cpue > 0) %>% 
  concaveman::concaveman(., concavity = 3) %>% #Draw concave hull around positive catches
  st_buffer(., dist = 45) %>%  # extend buffer
  st_difference(ak_land) #remove sections of polygon intersecting land

# intersect CPUE observations with buffer object
hair_cpue_mf <- hair_cpue_sf_mf %>%
  mutate(district = ifelse(district %in% c("ALL", "NORTH"), 
                           "N. EBS/NBS",
                           district)) %>% 
  st_intersection(., hair_loc_buffers_mf) %>% 
  sfc_as_cols(., names = c("longitude", "latitude")) %>% 
  st_set_geometry(NULL) %>% 
  as.data.frame()

# define new mesh
tinyVAST_mesh2 <- sdmTMB::make_mesh(hair_cpue_mf,
                                    xy_cols = c("longitude","latitude"),
                                    type = "kmeans",
                                    n_knots = 150)$mesh

ggplot() +
  geom_sf(data = hair_loc_buffers_mf, fill = "lightblue") + # buffer region
  geom_sf(data = hair_cpue_sf_mf, color = "grey") + # all observations
  geom_point(data = hair_cpue_mf %>% filter(cpue == 0), aes(x = longitude,
                                                             y = latitude), color = "red") +
  geom_point(data = hair_cpue_mf  %>% filter(cpue > 0), aes(x = longitude,
                                                             y = latitude)) +
  facet_wrap(~var)
```

#### Fitting the model

Here we specify that we want to model the effects of lag-1 male and female hair crab density on densities at time *t* We also model the cross-lagged effect of lag-1 mature female crab density on legal male density, and lag-1 mature male crab density on female crab density. Lastly, we specify an iid spatiotemporal random field component for each sex. For the sake of runtime, we exclude the spatial random field and anisotropy from the model.

```{r}

spacetime_term_vast <- "
  # autoregressive effects
  legal_male -> legal_male, 1, b11
  mature_female -> mature_female, 1, b22
  
  # cross lagged effects
  mature_female -> legal_male, 1, b21
  legal_male -> mature_female, 1, b12
  
  # spatiotemporal variances (seperate)
  legal_male <-> legal_male, 0, sd_spacetime1
  mature_female <-> mature_female, 0, sd_spacetime2
"
process <- FALSE
if (process){
  
  
    # fit model
  m2_tv <- tinyVAST(
             spacetime_term = spacetime_term_vast,
             data = hair_cpue_mf,
             time_column = "year",
             space_columns = c("longitude","latitude"),
             variable_column = "var",
             formula = cpue ~ 0 + var,
             spatial_domain = tinyVAST_mesh2,
             family = tweedie() )
  
  
  save(m2_tv, file = here::here("data/tinyVAST_m2.rdata"))
} else {
  load(here::here("data/tinyVAST_m2.rdata"))
}

```

#### Generating the prediction grid

We will use a slightly different prediction grid than before, but with the same spatial resolution. In this case, we intersect the grid with the male/female buffer instead of the male only buffer. We also additionally expand the prediction grid by a `var` variable indicating "legal_male" or "mature_female" crab CPUE.

```{r}
# create EBS grid for male/female crab
ebs_grid_mf <- ebs_domain %>% 
  st_make_grid(., cellsize  = cellsize) %>% 
  st_intersection(., hair_loc_buffers_mf) %>%
  st_as_sf() %>% 
  mutate(id = 1:nrow(.),
         area = as.numeric(st_area(.)))

# area vector to pass to integrate_output()
area_vec_mf <- ebs_grid_mf$area

# expand over time
pred_grid_mf <- ebs_grid_mf %>% 
  tidyr::expand_grid(.,
                     year = min(unique(hair_cpue_mf$year)):max(unique(hair_cpue_mf$year)),
                     var = c("legal_male", "mature_female")) %>% 
  st_as_sf() %>% 
  filter(year != 2020) # COVID year - no survey so should not be in final data

# final outputs
pred_df_mf <- pred_grid_mf %>% 
  st_centroid() %>% 
  sfc_as_cols(., names = c("longitude","latitude")) %>% 
  st_set_geometry(NULL) %>% 
  as.data.frame()
```

#### Predicting from the model

Predicting from `tinyVAST` models is similar to predicting from `mgcv` models. Where `sdmTMB` returned a `tibble`, `tinyVAST` returns a list of vectors. Predict from the model with `se.fit = FALSE` if you need to move quickly.

```{r, fig.width=8.5, fig.height=11}
if (process){
  m2_tv_predictions <- predict(m2_tv, 
                          newdata = pred_df_mf, se.fit = TRUE)
  save(m2_tv_predictions, file = here::here("data/vast_model_predictions.rdata"))
} else {
  load(here::here("data/vast_model_predictions.rdata")) 
}

m2_tv_predictions_df <- pred_df_mf %>% 
  mutate(est_tv = m2_tv_predictions$fit,
         se_tv = m2_tv_predictions$se.fit)

model_predictions_sf_mf <- pred_grid_mf %>% 
  left_join(., m2_tv_predictions_df, by = c("id","year","var")) 

ystart <- 2011
yend <- 2015
ggplot() +
  geom_sf(data = ak_land, fill = "darkgreen", alpha = 0.75) +
  geom_sf(data = model_predictions_sf_mf %>% 
            filter(between(year, ystart, yend)), aes(fill = log(est_tv),
                                                     color = log(est_tv))) +
  geom_point(data = hair_cpue_mf %>% filter(cpue > 0,
                                              between(year, ystart, yend)),
          aes(x = longitude, y = latitude,
              size = log(cpue)), color = "red", alpha = 0.25) +
  scale_fill_gradient2(low = scales::muted("red"),
                       high = scales::muted("blue"))+
  scale_color_gradient2(low = scales::muted("red"),
                       high = scales::muted("blue"))+
  guides(color = "none") +
  labs(fill = "Predicted log(CPUE)",
       size = "Observed log(CPUE>0)") +
  theme_light() + 
  coord_sf(ylim = c(650, 1500),
         xlim = c(-1300, -175)) +
  facet_grid(year ~ var) +
  theme(legend.position = "bottom")
```

#### DHARMa residuals

`tinyVAST` currently has functionality to visualize `DHARMa` residuals associated with fitted models (in addition to deviance and response residuals via `residuals()`). These tend to break on my machine when rendering the document so I save/load them here.

```{r, eval=FALSE}
if (process){
  new_data <- simulate(m2_tv, nsim = 100,type = "mle-mvn")
  save(new_data, file = here::here("data/m2_tv_dharma_sims.rdata"))
} else {
  load(here::here("data/m2_tv_dharma_sims.rdata"))
}
fitted <- predict(m2_tv)
res = DHARMa::createDHARMa( simulatedResponse = new_data, 
                            observedResponse = hair_cpue_mf$cpue, 
                            fittedPredictedResponse = fitted )
plot(res)
```

### Multiple response distributions

Lastly, we extend our bivariate VAST model by treating male and female hair crab CPUE as being drawn from distinct Tweedie distributions. The model takes \~ 8 minutes to run on my machine, and it throws errors when checked using the `sanity_tv` function. Regardless, the Tweedie parameters, reported as `log_sigma` in the outputs, are virtually equivalent for male and female observations.

```{r}
fam_list <- list(
 legal_male = tweedie(),
 mature_female = tweedie()
)

if (process){
  
  fam_list <- list(
    legal_male = tweedie(),
    mature_female = tweedie()
  )
  
  m3_tv <- tinyVAST(
           spacetime_term = spacetime_term_vast,
           data = hair_cpue_mf,
           time_column = "year",
           space_columns = c("longitude","latitude"),
           variable_column = "var",
           distribution_column = "var",  # must be supplied with >1 family arg
           formula = cpue ~ 0 + var,
           spatial_domain = tinyVAST_mesh2,
           family = fam_list )
  
  
  save(m3_tv, file = here::here("data/diff_tweeds.rdata"))
} else {
  load(here::here("data/diff_tweeds.rdata"))
}
```
